A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-05-02 paper.Rmd
abstract revised

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the existence of numerous hypothesis test procedures. This paper provides evidence for why this is good advice, using data from a visual inference experiment. We show how conventional tests are too sensitive to be practically useful, and how the lineup protocol can be used to yield reliable and consistent reading of residual plots for better model diagnosis. 
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also serves as an omnibus test, capable of detecting a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

### 2023-05-02 paper.tex
abstract revised

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the existence of numerous hypothesis test procedures.
This paper provides evidence for why this is good advice, using data
from a visual inference experiment. We show how conventional tests are
too sensitive to be practically useful, and how the lineup protocol can
be used to yield reliable and consistent reading of residual plots for
better model diagnosis.
\end{abstract}

@@

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model is deficient. It does not take into account the possible
dependencies in the visual test due to repeated evaluations of the same
lineup, or account for when participants are offered the option to
select one or more ``most different'' plots, or none, from a lineup.
They suggest three common lineup scenarios: (1) \(K\) different lineups
are shown to \(K\) participants, (2) \(K\) lineups with different null
plots but the same data plot are shown to \(K\) participants, and (3)
the same lineup is shown to \(K\) participants. Scenario 3 is the most
feasible to apply, but has the most dependencies to accommodate for the
\(p\)-value calculation. For Scenario 3, VanderPlas et al propose
modeling the probability of plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). The number of times plot \(i\) being
selected in \(K\) evaluations is denoted as \(c_i\). In case participant
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots participant \(j\)
selected for \(j=1,...K\). This ensures \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the data plot \(i\), the marginal
model can be simplified to a beta-binomial model and thus the visual
\(p\)-value is given as

@@

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
calculation of effect size are provided in Section \ref{effect-size}
after the introduction of the simulation model used in our experiment.
The theoretical power of a test is sometimes not a trival solution, but
it can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

Three different sample sizes were used in our experiment:
\(n = 50, 100, 300\). Figure \ref{fig:different-n} shows the non-linear
``S'' shape for different sample sizes. We expect signal strength to
decline in the simulated data plots with smaller \(n\). A sample size of
300 is typically enough for structure to be visible in a scatter plot
reliably. Beyond 300, structure can be obscured by over-plotting, and a
different type of display that included transparency, binning or density
mapping, might be needed.

@@

These \texttt{R} packages are used for the work: \texttt{cli}
\citep{cli}, \texttt{curl} \citep{curl}, \texttt{dplyr} \citep{dplyr},
\texttt{ggplot2} \citep{ggplot2}, \texttt{jsonlite} \citep{jsonlite},
\texttt{lmtest} \citep{lmtest}, \texttt{mpoly} \citep{mpoly},
\texttt{progress} \citep{progress}, \texttt{tibble} \citep{tibble},
\texttt{ggmosaic} \citep{ggmosaic}, \texttt{purrr} \citep{purrr},
\texttt{tidyr} \citep{tidyr}, \texttt{readr} \citep{readr},
\texttt{stringr} \citep{stringr}, \texttt{here} \citep{here},
\texttt{kableExtra} \citep{kableextra}, \texttt{patchwork}
\citep{patchwork}, \texttt{rcartocolor} \citep{rcartocolor}. The study
website is powered by \texttt{PythonAnywhere} \citep{pythonanywhere} and
the \texttt{Flask} web framework \citep{flask}. The \texttt{jsPsych}
framework \citep{jspsych} is used to create behavioral experiments that
run in our study website.

>>>>

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the availability of many numerical hypothesis test
procedures designed to use residuals to assess problems with a model
fit. Here we provide evidence for why this is good advice using data
from a visual inference experiment. We show how conventional tests are
too sensitive, which means that too often the conclusion would be that
the model fit is inadequate. The experiment uses the lineup protocol
which puts a residual plot in the context of null plots. This helps
generate reliable and consistent reading of residual plots for better
model diagnosis. It can also help in an obverse situation where a
conventional test would fail to detect a problem with a model due to
contaminated data. The lineup protocol also serves as an omnibus test,
capable of detecting a range of departures from good residuals
simultaneously.
\end{abstract}

@@

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model is deficient. It does not take into account the possible
dependencies in the visual test due to repeated evaluations of the same
lineup, or account for when participants are offered the option to
select one or more ``most different'' plots, or none, from a lineup.
They suggest three common lineup scenarios: (1) \(K\) different lineups
are shown to \(K\) participants, (2) \(K\) lineups with different null
plots but the same data plot are shown to \(K\) participants, and (3)
the same lineup is shown to \(K\) participants. Scenario 3 is the most
feasible to apply, but has the most dependencies to accommodate for the
\(p\)-value calculation. For Scenario 3, VanderPlas et al propose
modelling the probability of plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). The number of times plot \(i\) being
selected in \(K\) evaluations is denoted as \(c_i\). In case participant
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots participant \(j\)
selected for \(j=1,...K\). This ensures \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the data plot \(i\), the marginal
model can be simplified to a beta-binomial model and thus the visual
\(p\)-value is given as

@@

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
calculation of effect size are provided in Section \ref{effect-size}
after the introduction of the simulation model used in our experiment.
The theoretical power of a test is sometimes not a trivial solution, but
it can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

Three different sample sizes were used in our experiment:
\(n = 50, 100, 300\). Figure \ref{fig:different-n} shows the non-linear
``S'' shape for different sample sizes. We expect signal strength to
decline in the simulated data plots with smaller \(n\). A sample size of
300 is typically enough for structure to be visible in a scatterplot
reliably. Beyond 300, structure can be obscured by over-plotting, and a
different type of display that included transparency, binning or density
mapping, might be needed.

@@

These \texttt{R} packages are used for the work: \texttt{cli}
\citep{cli}, \texttt{curl} \citep{curl}, \texttt{dplyr} \citep{dplyr},
\texttt{ggplot2} \citep{ggplot2}, \texttt{jsonlite} \citep{jsonlite},
\texttt{lmtest} \citep{lmtest}, \texttt{mpoly} \citep{mpoly},
\texttt{progress} \citep{progress}, \texttt{tibble} \citep{tibble},
\texttt{ggmosaic} \citep{ggmosaic}, \texttt{purrr} \citep{purrr},
\texttt{tidyr} \citep{tidyr}, \texttt{readr} \citep{readr},
\texttt{stringr} \citep{stringr}, \texttt{here} \citep{here},
\texttt{kableExtra} \citep{kableextra}, \texttt{patchwork}
\citep{patchwork}, \texttt{rcartocolor} \citep{rcartocolor}. The study
website is powered by \texttt{PythonAnywhere} \citep{pythonanywhere} and
the \texttt{Flask} web framework \citep{flask}. The \texttt{jsPsych}
framework \citep{jspsych} is used to create behavioural experiments that
run in our study website.
